Descriptive Complexity for Counting Complexity Classes

### 2018-12-10 LMCS/beyond.tex
added reviewer 2 to todo list

\begin{exa} \label{ex:count-path}
We would like to define an $\rqfo(\fo)$-formula that, given a directed acyclic graph $G$ with $n$ nodes and a pair of nodes $b$, $c$ in $G$, counts the number of paths of length at most $n$ from $b$ to $c$ in $G$. To this end, assume that graphs are encoded using the signature $\R = \{ E(\cdot,\cdot)\}$, and then define formula $\alpha(x, y, f)$ as follows:
$$ 
E(x,y) + \sa{z} f(x,z)\cdot E(z,y).
$$
We have that $\clfp{\alpha(x,y,f)}$ defines our counting function. In fact, assume that $\A$ is an $\R$-structure with $n$ elements in its domain encoding an acyclic directed graph. Moreover, assume that $b,c$ are elements of $\A$ and $v$ is a first-order assignment over $\A$ such that $v(x) = b$ and $v(y) = c$. Then we have that $\sem{\clfp{\alpha(x,y,f)}}(\A,v)$ is equal to the  number of paths in $\A$ from $b$ to $c$ of length at most $n$.

>>>>

\begin{exa} \label{ex:count-path}
We would like to define an $\rqfo(\fo)$-formula that, given a directed acyclic graph $G$ with $n$ nodes and a pair of nodes $b$, $c$ in $G$, counts the number of paths of length less than $n$ from $b$ to $c$ in $G$. To this end, assume that graphs are encoded using the signature $\R = \{ E(\cdot,\cdot)\}$, and then define formula $\alpha(x, y, f)$ as follows:
$$ 
E(x,y) + \sa{z} f(x,z)\cdot E(z,y).
$$
We have that $\clfp{\alpha(x,y,f)}$ defines our counting function. In fact, assume that $\A$ is an $\R$-structure with $n$ elements in its domain encoding an acyclic directed graph. Moreover, assume that $b,c$ are elements of $\A$ and $v$ is a first-order assignment over $\A$ such that $v(x) = b$ and $v(y) = c$. Then we have that $\sem{\clfp{\alpha(x,y,f)}}(\A,v)$ is equal to the  number of paths in $\A$ from $b$ to $c$ of length at most $n$.

### 2018-12-10 LMCS/proof-capture-fpspace.tex
added reviewer 2 to todo list

For condition (1), notice that each $\pfp$ formula can be evaluated in deterministic polynomial space, that the constant function $s$ can be trivially simulated in $\fpspace$ for each $s\in\nat$, and that $\fpspace$ is closed under exponential sum and multiplication. This suffices to show that the condition holds.

%To show how to evaluate a $\qso(\pfp)$-formula, we construct recursively a $\shpspace$-machine $M_{\alpha}$ for each $\qso(\pfp)$ formula $\alpha$ over a signature $\R$. This machine runs in non-deterministic polynomial space and, on input $(\A,v,V)$, accepts in $\sem{\alpha}(\A,v,V)$ of its non-deterministic paths for each $(\A,v,V) \in \ostr[\R]^*$. Suppose $\A$ has domain $A$. If $\alpha$ is a $\pfp$-formula $\varphi$, then the machine checks if $(\A,v,V)\models\varphi$ deterministically in polynomial space~\cite{L04}, and accepts if and only if it holds true. If $\alpha$ is a constant $s$, it produces $s$ branches and accepts in all of them. If $\alpha = (\beta \add \gamma)$, then it chooses between 0 or 1, if it is 0 (1), it simulates $M_{\beta}$ ($M_{\gamma}$) on input $(\A,v,V)$. If $\alpha = (\beta \mult \gamma)$, it simulates $M_{\beta}$ on input $(\A,v,V)$ and on each accepting path, it continues simulating $M_{\gamma}$ on input $(\A,v,V)$.
%If $\alpha = \sa{x}\beta$, it chooses $a\in A$ non-deterministically and simulates $M_{\beta}$ on input $(\A,v[a/x],V)$. If $\alpha = \pa{x}\beta$, it simulates $M_{\beta}$ on input $(\A,v[a/x],V)$ consecutively for each $a\in A$, continuing to the next $A$-value if the run of $M_{\beta}$ accepts. If $\alpha = \sa{X}\beta$, it chooses $B\in A^{arity(X)}$ and simulates $M_{\beta}$ on input $(\A,v,V[B/X])$. If $\alpha = \pa{X}\beta$, it simulates $M_{\beta}$ on input $(\A,v,V[B/X])$ consecutively for each $B\in A^{arity(X)}$, again continuing to the next $A$-value if the run of $M_{\beta}$ accepts. This covers all possible cases for $\alpha$, and each of these steps can be computed in polynomial space. Let $\alpha$ be a formula in $\qso(\pfp)$ over a signature $\R$ and let $f$ be a function over $\R$ such that $f(\enc(\A))$ is equal to the accepting paths of $M_{\alpha}$ on input $(\A,v,V)$ for some $(\A,v,V) \in \ostr[\R]^*$. We have that $f$ is a $\shpspace$ function over $\R$, which implies that $f$ is also a $\fpspace$ function over $\R$, by the fact that $\shpspace = \fpspace$ \cite{Ladner89}, and that $f(\enc(\A)) = \sem{\alpha}(\A)$ for every $\A\in\ostr[\R]$.

For the condition (2), we provide a similar proof to theorem \ref{theo:capture-fp}. Let $f\in \fpspace$ defined over some $\R$. Let $\ell\in\nat$ be such that for each $\A\in\ostr[\R]$, $\lceil\log_2 f(\enc(\A)) \rceil \leq 2^{n^\ell}$ (i.e. $2^{n^\ell}$ is an upper bound for the output size), where $\A$ has a domain of size $n$. Let $X$ be a second-order variable of arity $\ell$. Consider a linear order over predicates of arity $\ell$ given by the formula 
$$
\varphi_{<}(X,Y) = \exists\bar{u}\big[\neg X(\bar{u})\wedge Y(\bar{u})\wedge \forall\bar{v}\big(
\bar{u}<\bar{v}\to(X(\bar{u})\iff Y(\bar{v}))\big)\big].
$$
Namely, we use relations to encode numbers with at most $2^{n^\ell}$-bits where the empty relation represents $0$ and the total-relation represents $2^{2^{n^\ell}}-1$.
Furthermore, each relation $X$ indexes a position in the binary output of $f(\enc(\A))$ as follows.
%Consider a polynomial space machine over the $\R$ that receives as input an $\R$-structure $\A$ and a number $p$ encoded by a relation $X$. Then the machine accepts if, and only if, the $p$-th bit of $f(\enc(\A))$ is $1$. 
Define a language
\[
L = \{(\A,B)\mid B \subseteq A^{\ell}\text{ and the $B$-th bit of $f(\enc(\A))$ is 1}\}.
\]
Since this language is in $\pspace$, it can be described in $\pfp$ \cite{AbiteboulV89} by a formula $\Phi(X)$ where the free variable $X$ encodes the number $p$. Then, similar than the previous proof we define:
$$
\alpha := \sa{X}(\Phi(X)\mult\varphi(X)),
$$ 
where $\varphi(X) = \pa{Y}(\varphi_{<}(Y,X)\mapsto 2)$. As in theorem \ref{theo:capture-fp}, this formula takes the value $2^m$ if there exist $m$ predicates that are smaller than $X$. Using an analogous argument, we conclude that $\alpha\in\qso(\pfp)$ and $\sem{\alpha}(\A) = f(\enc(\A))$.

>>>>

For the first condition of Definition~\ref{def:cap}, notice that each $\pfp$ formula can be evaluated in deterministic polynomial space, the constant function $s$ can be trivially simulated in $\fpspace$, and $\fpspace$ is closed under exponential sum and multiplication. This suffices to show that the condition holds.
For the second condition, the proof is similar than in Theorem~\ref{theo:capture-fp}. Let $f\in \fpspace$ defined over some $\R$ and $\ell\in\nat$ such that $\log_2\left( f(\enc(\A)) \right) \leq 2^{{|\A|}^\ell}$ for every $\A\in\ostr[\R]$  (i.e. $2^{{|\A|}^\ell}$ is an upper bound for the output size). Let $X$ be a second-order variable of arity $\ell$. Consider the linear order induced by $<$ over predicates of arity $\ell$ which can be defined by the following formula:
$$
\varphi_{<}(X,Y) = \ex{\bar{u}}\big[\neg X(\bar{u})\wedge Y(\bar{u})\wedge \fa{\bar{v}}\big(
\bar{u}<\bar{v}\to(X(\bar{u})\iff Y(\bar{v}))\big)\big].
$$
Namely, we use relations to encode numbers with at most $2^{{|\A|}^\ell}$ bits where the empty relation represents $0$ and the total-relation represents $2^{2^{{|\A|}^\ell}}-1$.
Furthermore, we can use a relation~$X$ to index a position in the binary output of $f(\enc(\A))$ as follows.
%Consider a polynomial space machine over the $\R$ that receives as input an $\R$-structure $\A$ and a number $p$ encoded by a relation $X$. Then the machine accepts if, and only if, the $p$-th bit of $f(\enc(\A))$ is $1$. 
Define the language:
\[
L = \{(\A,B)\mid B \subseteq A^{\ell}\text{ and the $B$-th bit of $f(\enc(\A))$ is 1}\}.
\]
Since $L$ is in $\pspace$, it can be specified in $\pfp$ \cite{AbiteboulV89} by a formula $\Phi(X)$ where the free variable $X$ encodes relation $B$ in $L$. Then, similar than the previous proof we define:
$$
\alpha := \sa{X} \Phi(X)\mult  \pa{Y}(\varphi_{<}(Y,X)\mapsto 2).
$$ 
where $\pa{Y}(\varphi_{<}(Y,X)\mapsto 2)$ takes the value $2^m$ if there exist $m$ predicates that are smaller than~$X$ and $\alpha$ reconstruct the output of $f(\enc(\A))$ by simulating $f$ with $\Phi(X)$. Using an analogous argument, we conclude that $\alpha\in\qso(\pfp)$ and $\sem{\alpha}(\A) = f(\enc(\A))$.

### 2018-12-10 LMCS/proof-sigma2hard.tex
added reviewer 2 to todo list

Next, we prove that $\shdhsat$ is hard for $\eqso(\ehorn)$ over a signature~$\R$ under parsimonious reductions. For each $\eqso(\ehorn)$ formula $\alpha$ over $\R$, we will define a polynomial-time function $g_{\alpha}$ that receives an $\R$-structure $\A$ and outputs an instance of $\shdhsat$ such that $\sem{\alpha}(\A) = \shdhsat(g_{\alpha}(\A))$. By Proposition~\ref{theo-pnf-snf}, we can assume that $\alpha$ is of the form:
$$
\alpha \ = \ \sum_{i = 1}^{m} \sa{\bar{X}_{i}}\sa{\bar{x}} \ex{\bar{y}} \bigwedge_{j = 1}^{n} \fa{\bar{z}} \varphi^i_j(\bar{X}_{i},\bar{x},\bar{y},\bar{z}),
$$
where each $\varphi^i_j$ is a Horn-clause, and each $\bar{X}_{i}$ is a sequence of second-order variables.
Consider $\bar{X}$ as the union of all $\bar{X}_{i}$. We replace each of the $m$ addends in $\alpha$ for
$$
\sa{\bar{X}}\sa{\bar{x}} \ex{\bar{y}} \bigwedge_{j = 1}^{n} \fa{\bar{z}} \varphi^i_j(\bar{X}_{i},\bar{x},\bar{y},\bar{z})\wedge\bigwedge_{X\not\in\bar{X}_{i}}\fa{\bar{u}}X(\bar{u}),
$$
whose sum is equivalent to $\alpha$.
Now, consider a finite $\R$-structure $\A$ with domain $A$. 
The next transformation of $\alpha$ and $\A$ towards a disjunction of Horn-formulas is to expand each first-order quantifier (i.e. $\Sigma{\bar{x}}$,  $\exists\bar{y}$, and $\forall\bar{z}$) as we replace their variables with first-order constants. Specifically, we obtain the following formula which defines the same function as $\alpha$ and is of polynomial size with respect to $\A$:
$$
\alpha_{\A} \ = \ \sum_{i = 1}^{m} \sum_{\bar{a}\,\in A^{\length{\bar{x}}}} \sa{\bar{X}}\bigvee_{\bar{b}\,\in A^{\length{\bar{y}}}}\bigwedge_{j = 1}^{n'}\bigwedge_{\bar{c}\,\in A^{\length{\bar{z}}}}\varphi^i_j(\bar{X},\bar{a},\bar{b},\bar{c}).
$$
Note that each first-order sub-formula in $\varphi^i_j(\bar{X},\bar{a},\bar{b},\bar{c})$ has no free variables and, therefore, we can evaluate each of them in polynomial time and replace by $\perp$ and $\top$ wherever it corresponds. In other words, in polynomial time we can replace $\varphi^i_j$ by a disjunction of $\neg X_{h}$ and at most one $X_{h}$, evaluated on constants.
After simplifying, grouping and reordering the previous formula, we can obtain an equivalent formula $\alpha_{\A}'$ of the form:
$$
\alpha_{\A}' \ := \ \sum_{i = 1}^{m'}\sa{\bar{X}}\bigvee_{j = 1}^{n_1'}\bigwedge_{k = 1}^{n_2'}\psi^{i}_{j,k}(\bar{X})
$$
where every $\psi^{i}_{j,k}(\bar{X})$ is a disjunction of $\neg X_{h}$ and at most one $X_{h}$.
The reader can easily check that $\sem{\alpha}(\A) = \sem{\alpha_{\A}'}(\A)$.

>>>>

Next, we prove that $\shdhsat$ is hard for $\eqso(\ehorn)$ over a signature~$\R$ under parsimonious reductions. For each $\eqso(\ehorn)$ formula $\alpha$ over $\R$, we will define a polynomial-time function $g_{\alpha}$ that receives an $\R$-structure $\A$ and outputs an instance of $\shdhsat$ such that $\sem{\alpha}(\A) = \shdhsat(g_{\alpha}(\A))$. By Proposition~\ref{theo-pnf-snf}, we can assume that $\alpha$ is of the form:
$$
\alpha \ = \ \sum_{i = 1}^{m} \sa{\bar{X}_{i}}\sa{\bar{x}} \ex{\bar{y}} \bigwedge_{j = 1}^{n} \fa{\bar{z}} \varphi^i_j(\bar{X}_{i},\bar{x},\bar{y},\bar{z}),
$$
where each $\varphi^i_j$ is a Horn-clause, and each $\bar{X}_{i}$ is a sequence of second-order variables.
Consider $\bar{X}$ as the union of all $\bar{X}_{i}$. We replace each of the $m$ addends in $\alpha$ with
$$
\sa{\bar{X}}\sa{\bar{x}} \ex{\bar{y}} \bigwedge_{j = 1}^{n} \fa{\bar{z}} \varphi^i_j(\bar{X}_{i},\bar{x},\bar{y},\bar{z})\wedge\bigwedge_{X\not\in\bar{X}_{i}}\fa{\bar{u}}X(\bar{u}),
$$
whose sum is equivalent to $\alpha$.
Now, consider a finite $\R$-structure $\A$ with domain $A$. 
The next transformation of $\alpha$ and $\A$ towards a disjunction of Horn-formulas is to expand each first-order quantifier (i.e. $\Sigma{\bar{x}}$,  $\exists\bar{y}$, and $\forall\bar{z}$) as we replace their variables with first-order constants. Specifically, we obtain the following formula which defines the same function as $\alpha$ and is of polynomial size with respect to $\A$:
$$
\alpha_{\A} \ = \ \sum_{i = 1}^{m} \sum_{\bar{a}\,\in A^{\length{\bar{x}}}} \sa{\bar{X}}\bigvee_{\bar{b}\,\in A^{\length{\bar{y}}}}\bigwedge_{j = 1}^{n'}\bigwedge_{\bar{c}\,\in A^{\length{\bar{z}}}}\varphi^i_j(\bar{X},\bar{a},\bar{b},\bar{c}).
$$
Note that each first-order sub-formula in $\varphi^i_j(\bar{X},\bar{a},\bar{b},\bar{c})$ has no free variables and, therefore, we can evaluate each of them in polynomial time and replace with $\perp$ and $\top$ wherever it corresponds. In other words, in polynomial time we can replace $\varphi^i_j$ with a disjunction of $\neg X_{h}$ and at most one $X_{h}$, evaluated on constants.
After simplifying, grouping and reordering the previous formula, we can obtain an equivalent formula $\alpha_{\A}'$ of the form:
$$
\alpha_{\A}' \ := \ \sum_{i = 1}^{m'}\sa{\bar{X}}\bigvee_{j = 1}^{n_1'}\bigwedge_{k = 1}^{n_2'}\psi^{i}_{j,k}(\bar{X})
$$
where every $\psi^{i}_{j,k}(\bar{X})$ is a disjunction of $\neg X_{h}$ and at most one $X_{h}$.
The reader can easily check that $\sem{\alpha}(\A) = \sem{\alpha_{\A}'}(\A)$.

### 2018-12-10 LMCS/proof-sigmafo-minusone.tex
added reviewer 2 to todo list

\noindent {\bf Closed under subtraction by one.} We prove here that $\eqso(\logex{1})$ is closed under subtraction by one. 
For this, given $\alpha \in \eqso(\logex{1})$ over a signature $\R$, we will define a $\eqso(\logex{1})$-formula $\kappa(\alpha)$ such that for each finite structure $A$ over $\R$: $\sem{\kappa(\alpha)}(\A) = \sem{\alpha}(\A) \dotminus 1$. 
Without loss of generality, we assume that $\alpha$ is in $\logex{1}$-SNF, that is, $\alpha = \sum_{i = 1}^{n}\sa{\bar{X}}\sa{\bar{x}}\varphi_i$ where each $\varphi_i$ is in $\logex{1}$. Moreover, we assume that $\length{\bar{x}} > 0$  since, if this is not the case, we can replace $\sa{\bar{X}}\varphi_i$ for the equivalent formula $\sa{\bar{X}} \sa{y} \varphi_i \wedge\first(y)$.

@@

Now, we have all the ingredients in order to define $\kappa(\alpha)$. 
Intuitively, we want to exclude the assignment $(V^*, v^*)$ from the satisfying assignments of $\varphi_i(\bar{X},\bar{x})$.
Towards this goal, we can define a formula $\psi_i(\bar{X},\bar{x})$ such that $(\A, V, v) \models \psi_i(\bar{X},\bar{x})$ if, and only if, if $\varphi_i(\bar{X}, \bar{x})$ is satisfiable, then
$V \neq V^*$ or $v \neq v^*$. 
This property can be defined as follows:
\begin{align}
\psi_i(\bar{X},\bar{x}) \; := \;  &\big( \, \ex{\bar{x}} \ex{\bar{y}} \varphi^{\fo}_i(\bar{x},\bar{y}) \,\big) \rightarrow  \label{eq:part1} \\
&\Big( \; 
\ex{\bar{y}} \text{$min$-}\varphi^{\fo}_i(\bar{x}, \bar{y}) \wedge \big( \; \varphi_i'(\bar{X},\bar{x}, \bar{y})  \rightarrow \!\! \bigvee_{X \in \bar{X}} \ex{\bar{z}} [ \, X(\bar{z}) \wedge   \!\!\!\!\!\!\!\!\!\! \bigwedge\limits_{X(\bar{w}) \, \in \, \varphi^{+}_i(\bar{X},\bar{y})} \!\!\!\!\!\!\!\!\!\!\! \bar{w}\neq\bar{z} \, ] \; \big) \;  \vee   \label{eq:part2} \\ 
&(\; \ex{\bar{x}'} \ex{\bar{y}} \varphi_i'(\bar{X},\bar{x}',\bar{y})\wedge \bar{x}' < \bar{x} \; ) \;\; \Big)  \label{eq:part3}
\end{align}
To understand the formula, first notice that the premise of the implication at (\ref{eq:part1}) is true if, and only if, $\varphi_i(\bar{X}, \bar{x})$ is satisfiable. 
Indeed, by Claim~\ref{claim:minusone} we know that if $\ex{\bar{x}} \ex{\bar{y}} \varphi^{\fo}_i(\bar{x},\bar{y})$ is true, then there exists assignments $V$ and $v$ such that $(\A,V,v) \models\varphi_i'(\bar{X},\bar{x},\bar{y})$.
The conclusion of the implication (divided into (\ref{eq:part2}) and (\ref{eq:part3})), take care that $V(\bar{X}) \neq V^*(\bar{X})$ or $v(\bar{x}) \neq v^*(\bar{x})$.
Here, the first disjunct (\ref{eq:part2}) checks that $V(\bar{X}) \neq V^*(\bar{X})$ by defining that if $\varphi_i'(\bar{X},\bar{x}, \bar{y})$ is satisfied then $V^*(\bar{X}) \subsetneq V(\bar{X})$. 
The second disjunct (\ref{eq:part3}) is satisfied when $v(\bar{x})$ is not the lexicographically smallest tuple that satisfies $\varphi_i$ (i.e. $v(\bar{x}) \neq v^*(\bar{x})$).
Finally, from the previous discussion one can easily check that $\psi_i(\bar{X},\bar{x})$ satisfies the desire property.

>>>>

\noindent {\bf Closed under subtraction by one.} We prove here that $\eqso(\logex{1})$ is closed under subtraction by one. 
For this, given $\alpha \in \eqso(\logex{1})$ over a signature $\R$, we will define a $\eqso(\logex{1})$-formula $\kappa(\alpha)$ such that for each finite structure $A$ over $\R$: $\sem{\kappa(\alpha)}(\A) = \sem{\alpha}(\A) \dotminus 1$. 
Without loss of generality, we assume that $\alpha$ is in $\logex{1}$-SNF, that is, $\alpha = \sum_{i = 1}^{n}\sa{\bar{X}}\sa{\bar{x}}\varphi_i$ where each $\varphi_i$ is in $\logex{1}$. Moreover, we assume that $\length{\bar{x}} > 0$  since, if this is not the case, we can replace $\sa{\bar{X}}\varphi_i$ with the equivalent formula $\sa{\bar{X}} \sa{y} \varphi_i \wedge\first(y)$.

@@

Now, we have all the ingredients in order to define $\kappa(\alpha)$. 
Intuitively, we want to exclude the assignment $(V^*, v^*)$ from the satisfying assignments of $\varphi_i(\bar{X},\bar{x})$.
Towards this goal, we can define a formula $\psi_i(\bar{X},\bar{x})$ such that, if $\varphi_i(\bar{X}, \bar{x})$ is satisfiable, $(\A, V, v) \models \psi_i(\bar{X},\bar{x})$ if, and only if either $V \neq V^*$ or $v \neq v^*$. 
This property can be defined as follows:
\begin{align}
\psi_i(\bar{X},\bar{x}) \; := \;  &\big( \, \ex{\bar{x}} \ex{\bar{y}} \varphi^{\fo}_i(\bar{x},\bar{y}) \,\big) \rightarrow  \label{eq:part1} \\
&\Big( \; 
\ex{\bar{y}} \text{$min$-}\varphi^{\fo}_i(\bar{x}, \bar{y}) \wedge \big( \; \varphi_i'(\bar{X},\bar{x}, \bar{y})  \rightarrow \!\! \bigvee_{X \in \bar{X}} \ex{\bar{z}} [ \, X(\bar{z}) \wedge   \!\!\!\!\!\!\!\!\!\! \bigwedge\limits_{X(\bar{w}) \, \in \, \varphi^{+}_i(\bar{X},\bar{y})} \!\!\!\!\!\!\!\!\!\!\! \bar{w}\neq\bar{z} \, ] \; \big) \;  \vee   \label{eq:part2} \\ 
&(\; \ex{\bar{x}'} \ex{\bar{y}} \varphi_i'(\bar{X},\bar{x}',\bar{y})\wedge \bar{x}' < \bar{x} \; ) \;\; \Big)  \label{eq:part3}
\end{align}
To understand the formula, first notice that the premise of the implication at (\ref{eq:part1}) is true if, and only if, $\varphi_i(\bar{X}, \bar{x})$ is satisfiable. 
Indeed, by Claim~\ref{claim:minusone} we know that if $\ex{\bar{x}} \ex{\bar{y}} \varphi^{\fo}_i(\bar{x},\bar{y})$ is true, then there exists assignments $V$ and $v$ such that $(\A,V,v) \models\varphi_i'(\bar{X},\bar{x},\bar{y})$.
The conclusion of the implication (divided into (\ref{eq:part2}) and (\ref{eq:part3})), take care that $V(\bar{X}) \neq V^*(\bar{X})$ or $v(\bar{x}) \neq v^*(\bar{x})$.
Here, the first disjunct (\ref{eq:part2}) checks that $V(\bar{X}) \neq V^*(\bar{X})$ by defining that if $\varphi_i'(\bar{X},\bar{x}, \bar{y})$ is satisfied then $V^*(\bar{X}) \subsetneq V(\bar{X})$. 
The second disjunct (\ref{eq:part3}) is satisfied when $v(\bar{x})$ is not the lexicographically smallest tuple that satisfies $\varphi_i$ (i.e. $v(\bar{x}) \neq v^*(\bar{x})$).
Finally, from the previous discussion one can easily check that $\psi_i(\bar{X},\bar{x})$ satisfies the desire property.

### 2018-12-10 LMCS/proof-tqfo-shl.tex
added reviewer 2 to todo list

%&$\boldsymbol{\tqfo(\fo)}$ {\bf can be computed in} $\boldsymbol{\shl.}$
First, we show that every formula in $\tqfo(\fo)$ defines a function that is in $\shl$.
Let $\R$ be a relational signature and $\alpha$ a formula over $\R$ in $\tqfo(\fo)$. We construct a logarithmic-space nondeterministic Turing Machine $M_{\alpha}$ that on input $(\enc(\A),v)$, where $\A$ is an $\R$-structure and $v$ is a first-order assignment for $\A$, has $\sem{\alpha}(\A,v)$ accepting runs (so that we can conclude that the function defined by $\alpha$ is in $\shl$). Suppose that the domain of $\A$ is $A = \{1,\ldots,n\}$. The TM $M_{\alpha}$ needs $\ell \mult\log_2(n)$ bits of memory to store the first-order variables occurring in $\alpha$, where $\ell$ is the number of variables occurring in this formula (which is the same as the number of variables in the domain of $v$). If $\alpha = \varphi$, where $\varphi$ is an $\fo$-formula, then we check if $(\A,v)\models\varphi$ in deterministic logarithmic space, and accept if and only if this condition holds. If $\alpha = s$, where $s$ is a fixed natural number, then we generate $s$ possible runs and accept in all of them. If $\alpha = (\alpha_1 + \alpha_2)$, we simulate $M_{\alpha_1}$ and $M_{\alpha_2}$ on separate branches. If $\alpha = (\alpha_1\mult\alpha_2)$, we simulate $M_{\alpha_1}$ and if it accepts, then instead of accepting we simulate $M_{\alpha_2}$. If $\alpha = \sa{x}\beta$, for each $a\in A$ we generate a different run where we simulate $M_{\beta}$ with input $v[a/x]$. If $\alpha = \pa{x}\beta$, we simulate $M_{\beta}$ with input $v[1/x]$, and on each accepting run, instead of accepting we replace the assignment of $x$ to 2, to simulate $M_{\beta}$ with input $v[2/x]$, and so on. If $\alpha = [\pth \varphi(\bar{x},\bar{y})]$, where $\varphi$ is an $\fo$-formula, then we simulate the $\shl$ procedure that counts the number of paths of a given length from a source to a target node in an input graph (where the length is at most the number of nodes in the graph).
%This procedure starts by setting $\bar{a} = v(\bar{x})$. On each iteration, it nondeterministically chooses an assignment $\bar{b}$ for $\bar{x}$, and continues if $(\A,v)\models\varphi(\bar{a},\bar{b})$ where $\bar{a}$ is the previously chosen value for $\bar{x}$, and it rejects otherwise. If at any point we obtain that the current value for $\bar x$ is  $\bar{a}$ and $\bar a = v(\bar{y})$, we generate an accepting branch, and continue simulating the procedure in the current branch. We simulate $n^{\length{\bar{x}}}$ iterations of the procedure, and this generates exactly $\sem{[\pth \varphi(\bar{x},\bar{y})]}(\A,v)$ accepting branches. This ends the construction of the algorithm. Consider $f$ as the $\shl$ function associated to this procedure and we have that for each finite $\R$-structure $\A$: $f(\enc(\A)) = \sem{\alpha}(\A)$.

>>>>

%&$\boldsymbol{\tqfo(\fo)}$ {\bf can be computed in} $\boldsymbol{\shl.}$
First, we show that every formula in $\tqfo(\fo)$ defines a function that is in $\shl$.
Let $\R$ be a relational signature and $\alpha$ a formula over $\R$ in $\tqfo(\fo)$. We construct a logarithmic-space nondeterministic Turing Machine $M_{\alpha}$ that on input $(\enc(\A),v)$, where $\A$ is an $\R$-structure and $v$ is a first-order assignment for $\A$, has $\sem{\alpha}(\A,v)$ accepting runs (so that we can conclude that the function defined by $\alpha$ is in $\shl$). Suppose that the domain of $\A$ is $A = \{1,\ldots,n\}$. The TM $M_{\alpha}$ needs $\ell \mult\log_2(n)$ bits of memory to store the first-order variables that appear in $\alpha$, where $\ell$ is the number of different variables in this formula (and also the size of the domain of $v$). If $\alpha = \varphi$, where $\varphi$ is an $\fo$-formula, then we check if $(\A,v)\models\varphi$ in deterministic logarithmic space, and accept if and only if this condition holds. If $\alpha = s$, where $s$ is a fixed natural number, then we generate $s$ branches and accept in all of them. If $\alpha = (\alpha_1 + \alpha_2)$, we simulate $M_{\alpha_1}$ and $M_{\alpha_2}$ on separate branches. If $\alpha = (\alpha_1\mult\alpha_2)$, we simulate $M_{\alpha_1}$ and if it accepts, then instead of accepting we simulate $M_{\alpha_2}$. If $\alpha = \sa{x}\beta$, for each $a\in A$ we generate a different branch where we simulate $M_{\beta}$ with input $v[a/x]$. If $\alpha = \pa{x}\beta$, we simulate $M_{\beta}$ with input $v[1/x]$, and on each accepting run, instead of accepting we simulate $M_{\beta}$ with input $v[2/x]$, and so on. If $\alpha = [\pth \varphi(\bar{x},\bar{y})]$, where $\varphi$ is an $\fo$-formula, we simulate the $\shl$ procedure that counts the number of paths of a given length from a source to a target node in an input graph (where the length is at most the number of nodes in the graph).
%This procedure starts by setting $\bar{a} = v(\bar{x})$. On each iteration, it nondeterministically chooses an assignment $\bar{b}$ for $\bar{x}$, and continues if $(\A,v)\models\varphi(\bar{a},\bar{b})$ where $\bar{a}$ is the previously chosen value for $\bar{x}$, and it rejects otherwise. If at any point we obtain that the current value for $\bar x$ is  $\bar{a}$ and $\bar a = v(\bar{y})$, we generate an accepting branch, and continue simulating the procedure in the current branch. We simulate $n^{\length{\bar{x}}}$ iterations of the procedure, and this generates exactly $\sem{[\pth \varphi(\bar{x},\bar{y})]}(\A,v)$ accepting branches. This ends the construction of the algorithm. Consider $f$ as the $\shl$ function associated to this procedure and we have that for each finite $\R$-structure $\A$: $f(\enc(\A)) = \sem{\alpha}(\A)$.

### 2018-12-10 LMCS/todo-list.tex
added reviewer 2 to todo list

@@

	\item p.24: ``if, and only if, if X, then Y'': This construction makes it unneccesarily hard to parse

@@

	\item[$\checkmark$] p.37: ``we simulate $M_\beta$ with input $v[1/n]$'' $\to$ ``... with input $v[1/x]$''

@@

	\item on definition of functions, \texttt{\symbol{92}colon} should be used instead of :

@@

>>>>

{\bf Reviewer 1}

@@

	\item[$\checkmark$] p.24: ``if, and only if, if X, then Y'': This construction makes it unneccesarily hard to parse

@@

	\item[$\checkmark$] p.37: ``we simulate $M_\beta$ on input $v[1/n]$'' $\to$ ``... with input $v[1/x]$''

@@

	\item on definition of functions, \verb|\colon| should be used instead of :

@@

{\bf Reviewer 2}

\begin{itemize}
		\setlength\itemsep{0.5em}
\item In several proofs, you rely on disjoint union of solutions and cartesian product of solutions. Indeed, its gives a polynomial time counting reduction. Can you make connections with (+,*) circuits ? It would be similar to what is done in A Circuit-Based Approach to Efficient Enumeration (Antoine Amarilli, Pierre Bourhis, Louis Jachiet, Stefan Mengel) where simple circuits (dDNNF, which are basically disjoint unions and cartesian products) are used to capture the solutions of MSO queries.

\item In general I would be interested to know the relationship of your couting classes with their enumeration counterpart. 
For instance there is a paper adapting [39] to enumeration: Enumeration Complexity of Logical Query Problems with Second-order Variables (Arnaud Durand and Yann Strozecki). In particular they consider class with some Horn structure to capture easy enumeration problems, it may have some connection with the class $\sigma$ QSO($\Pi_1$ Horn) you introduce.

\item For the most involved proofs, you could describe informally your proof strategy in a few lines to help the reader. You may also add some example to illustrate the different fragment of QSO (you already do it sometimes and it is very helpful).

\item[$\checkmark$] page 4:
-some of the $\to$ some of them 
\item[$\checkmark$] page 10: 
-$2^{|A|^l}$-bits $\to$ $2^{|A|^l}$ bits
\item[$\checkmark$] page 21:
-it does not mentioned $\to$ it does not mention
\item page 22: 
-you make a long proof about some closure under subtraction by one.
You should put the result in a theorem envirpnment to make what you are
proving more evident.
\item[$\checkmark$] -replace .... for $\to$ replace ... by;  there are several instances of
that
\item[$\checkmark$] -$\phi$ is an FO-formula $\to$ a $\Sigma_1[FO]$ formula
\item -you explain that you rewrite your formula to obtain conditions a,b and c, 
you should give an idea how these conditions will be used
\item[$\checkmark$] page 23:
a SO, a FO  an SO, an FO
\item page 24: 
if, if $\phi$$\to$ $\phi$
\item page 25: 
-you explain that the proof relies on witnesses of logarithmic size.
It was not clear to me while reading the proof. Can you explain more ?
\item page 28:
-the complete problem for $\Sigma$ QSO($\Sigma_2$ Horn) would be complete also for $\#\Sigma-2$ Horn. You should mention here that you do not need sum here
and that the framework of [39] is enough for this result.
\item[?] page 30:
-not, trivial $\to$ not trivial
\item[$\checkmark$] page 32: 
- paths of length at most n $\to$ in an acylic graph, all paths are of size less than n
\item page 33:
-there is a proof symbol \verb|\qed| but it does not end a well defined proof environment

\item page 38: 
the last sentence on fixed point operator is to vague. Either expand it to 
say what king of question on fixed point operator you want to investigate or remove it	
\end{itemize}
